Error Reporting Parsers: a Monad Transformer Approach

### 2013-07-08 article.tex
remove resolved comments

\EZY{An alternate structure you can use here is to introduce Woof first, before talking about monad transformers.  
After all, the example parser doesn't require you to know what is going on in transformer-land.}

@@

While our previous parsers had two different results -- Nothing and Just --
our new parsers have three possible results:  success, failure, and error. 
A parse failure means that a parser wasn't able to match the input, but
nothing bad happened; failures can be recovered from if there is an alternative
parse that succeeds.  However, an error means that an unrecoverable problem
has been found and that parsing should immediately stop; the parser will
not be able to recover, even if there are alternatives since backtracking
will not be performed.
\EZY{Question a reader might have: what is the difference between failure and error?  
You give a little guidance later but give some intuition now!}
Note that successful and failing parses are wrapped in an additional \verb+Right+
constructor:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

@@

This combinator takes two arguments: an error value and a monadic 
computation.  It first tries to run the computation; if the computation 
fails, it generates an error.  The 
definition of 'failure' is provided by the Alternative instance of \verb+m+.
\EZY{Give a little intuition on why the operator is called commit earlier;
you don't get around it until later.}
The reason for its name is that it restricts backtracking and alternatives;
it ``commits" the parser to a do-or-die situation -- either it successfully
parses the input, or the parsing stops.  It's based on the \verb+nofail+
combinator of \cite{hutton1992higher}.

@@

Note that \verb+commit+ isn't used until we can be sure that we're in the right 
rule: \EZY{Here it is}  if, in the \verb+define+ parser, we committed to parsing the ``define" 
keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
that failed.  Thus, if errors are treated as unrecoverable, it's important not to 
place a \verb+commit+ where backtracking might be necessary to parse a valid 
alternative.  In our example, however, once 
we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule; allowing useless backtracking would only destroy our ability to report 
errors cleanly and accurately.

@@

\subsection{Position reporting}
To actually report the position when an error is detected, we introduce a new 
combinator, which I'll call \verb+cut+:
\Matt{a horrible name}
\begin{verbatim}
cut :: String -> Parser Char a -> Parser Char a
cut message parser = 
    getState >>= \p ->
    commit (message, p) parser
\end{verbatim}

@@

\Matt{hmm, the last point can be a disadvantage as well}

>>>>

@@

While our previous parsers had two different results -- Nothing and Just --
our new parsers have three possible results:  success, failure, and error. 
A parse failure means that a parser wasn't able to match the input, but
nothing bad happened; failures can be recovered from if there is an alternative
parse that succeeds.  However, an error means that an unrecoverable problem
has been found and that parsing should immediately stop; the parser will
not be able to recover, even if there are alternatives since backtracking
will not be performed.
Note that successful and failing parses are wrapped in an additional \verb+Right+
constructor:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

@@

This combinator takes two arguments: an error value and a monadic 
computation.  It first tries to run the computation; if the computation 
fails, it generates an error.  The 
definition of 'failure' is provided by the Alternative instance of \verb+m+.
The reason for its name is that it restricts backtracking and alternatives;
it ``commits" the parser to a do-or-die situation -- either it successfully
parses the input, or the parsing stops.  It's based on the \verb+nofail+
combinator of \cite{hutton1992higher}.

@@

Note that \verb+commit+ isn't used until we can be sure that we're in the right 
rule: if, in the \verb+define+ parser, we committed to parsing the ``define" 
keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
that failed.  Thus, if errors are treated as unrecoverable, it's important not to 
place a \verb+commit+ where backtracking might be necessary to parse a valid 
alternative.  In our example, however, once 
we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule; allowing useless backtracking would only destroy our ability to report 
errors cleanly and accurately.

@@

\subsection{Position reporting}
To actually report the position when an error is detected, we introduce a new 
combinator, which I'll call \verb+cut+:
\begin{verbatim}
cut :: String -> Parser Char a -> Parser Char a
cut message parser = 
    getState >>= \p ->
    commit (message, p) parser
\end{verbatim}

@@
